The origin of cosmic-ray electrons in cluster outskirts

### 2011-08-25 paper.tex
update text

\def\del#1{{}}
%\def\del#1{{\bf (DELETED TEXT)}}
\def\C#1{#1}
%\def\C#1{{\bf #1}}

@@

% --- Anders's commands --- %
\newcommand{\ev}{\mathrm{eV}}
\newcommand{\gev}{\mathrm{GeV}}
\newcommand{\mev}{\mathrm{MeV}}
\newcommand{\tev}{\mathrm{TeV}}
\newcommand{\mpc}{m_{\mathrm{p}}c}
\newcommand{\mpcc}{m_{\mathrm{p}}c^2}
\newcommand{\p}{\mathrm{p}}
\newcommand{\pp}{\mathrm{pp}}
\newcommand{\br}{\bold{r}}
\newcommand{\e}{\mathrm{e}}
\newcommand{\ii}{\mathrm{i}}
\newcommand{\inj}{\mathrm{inj}}
\newcommand{\pd}{\partial}
\newcommand{\CR}{\mathrm{CR}}
\newcommand{\gas}{\mathrm{gas}}
\newcommand{\mr}{\mathrm}
\newcommand{\DM}{\mathrm{DM}}
\newcommand{\ISM}{\mathrm{ISM}}
\newcommand{\ICM}{\mathrm{ICM}}
\newcommand{\CMB}{\mathrm{CMB}}
\newcommand{\IC}{\mathrm{IC}}
\newcommand{\mecc}{m_\e c^2}
\newcommand{\mug}{\mathrm{\mu G}}
\newcommand{\rvir}{R_{200}}
\newcommand{\B}{\mathrm{B}}
\newcommand{\ma}{\mathrm{max}}
\newcommand{\low}{\mathrm{low}}

@@

The CR electrons cool through inverse Compton (IC) emission and
Coulomb losses on timescales that are relative short compared to the
dynamical timescale of a cluster. We model these looses analytically
by assuming a self similar solution for the CR electron distribution
function \citep{1999ApJ...520..529S}
\begin{eqnarray}
  \label{eq:f_inj_e}
  f_\e(p_\e) &\approx& \frac{f_\rmn{inj,e}(\gamma_\e)}{\alpha-1}
\left(\frac{\gamma_\low}{\gamma_\e}+\frac{\gamma_\e }{\gamma_\ma}\right)^{-1}
\nonumber\\
&\times& \left\{2-\left[\frac{1}{2}\left(1-\frac{\gamma_\e}{\gamma_\ma}+
\left|1-\frac{\gamma_\e}{\gamma_\ma}\right|\right)\right]^{\alpha-1}\right.
\nonumber\\
&-& \left.\left(1+\frac{\gamma_\low}{\gamma_\e}\right)^{-(\alpha-1)}\right\}\,,
\end{eqnarray}
Here $\gamma_\e \approx p_\e = P_\e/m_\e c$ for $ p_\e \gg 1$.  The
approximation in Eqn.~(\ref{eq:f_inj_e}) is valid as long as
$\gamma_\ma\gg \gamma_\low$, where the upper limit is determined from
IC losses through
\begin{eqnarray}
  \label{eq:gamma_max}
b_\IC(\gamma_\e) &=& b_\IC \gamma_\e^2 
= \frac{4}{3}\frac{\sigma_\rmn{T}}{m_\e c} U_\rmn{cmb} \gamma_\e^2 \,,\quad \rmn{and}\nonumber \\
%1.37\times 10^{-20} \gamma^2(1+z)^4 \,\rmn{s}^{-1}
\gamma_\ma &=& \frac{1}{b_\IC\,\Delta t}\,.
\end{eqnarray}
Here $\sigma_\rmn{T}= 8\pi e^4/3(m_\e c^2)^2$ is the Thomson cross
section and $U_\rmn{cmb}$ is the energy density of the CMB in the
cluster. The lower limit is determined from Coulomb losses through
\begin{eqnarray}
  \label{eq:gamma_min}
b_\rmn{C}(\gamma_\e) &=& b_\rmn{C} \gamma_\e^2 
= \frac{3\,\sigma_\rmn{T}\,n_\rmn{el}\,c}{2}
\left[\ln\left(\frac{m_\e c^2 \sqrt{\gamma_\e-1}}{\hbar\,\omega_\rmn{plasma}}\right)\right.\nonumber \\
  &-&\ln(2)\left(\frac{1}{2}+\frac{1}{\gamma_\e}\right)+\frac{1}{2}+
\left.\left(\frac{\gamma_\e-1}{4\gamma_\e}\right)^2\right]
         \,,\quad \rmn{and}\nonumber \\
\gamma_\low &=& b_\rmn{C}\,\Delta t\,.
\end{eqnarray}
Here $\omega_\rmn{plasma} = \sqrt{4\pi e^2 n_\e / m_\e}$ is the plasma
frequency, and $n_\e$ is the number density of free electrons.

We model the final CR electron spectrum as a superposition of five CR
populations, each determined by Eqn.~\ref{eq:f_inj_e} but with a
different spectral index and injected spectra.

\subsection{Synchrotron emission}
In this section we derive the synchrotron emission from the CR
electron distribution.  WORK UNDER PROGRESS

@@

high energy part build up during last Gyr, also mention normalization
\begin{figure*}
\begin{minipage}{2.0\columnwidth}
  \includegraphics[width=0.49\columnwidth]{./figures/f_evolution.0.3Rv.eps}
  \includegraphics[width=0.49\columnwidth]{./figures/f_evolution.1.4Rv.eps}
  \caption{Time evolution of cosmic ray (CR) electron spectra. We show
    the CR electron distribution function weighted with the electron
    Lorentz factor ($\gamma_\e$) squared. The time evolution of the CR
    electrons which at $z=0$ reside from (0.3-0.5)$\rvir$ are shown in
    the {\it left panel} and (1.3-1.5)$\rvir$ in the {\it right
      panel}. The black lines show the CR electrons with a
    $\gamma_\e=30$, blue lines $\gamma_\e=200$, and green lines
    $\gamma_\e=10^4$. The solid lines show the CR electrons with full
    cooling, i.e. for Coulomb, inverse Compton, and adiabatic
    losses. Dashed lines show the CR electrons without Coulomb and
    inverse Compton cooling while the dash-dotted lines show the CR
    electrons without adiabatic cooling. The dotted line show the
    injected CR electron distribution function, derived from the
    injected CR protons. For each energy, we normalize the CR electron
    distribution function with the CR electrons without
    cooling. Notice that the high energy part is built up during the
    last Gyr. This is explained by the inverse Compton losses that are
    larger at high energies, especially at high redshifts where the
    energy density of the CMB is much larger than
    today.  \label{fig:e_evol}}
\end{minipage}
\end{figure*}

>>>>

\def\del#1{{}}
%\def\del#1{{\bf (DELETED TEXT)}}
%\def\C#1{#1}
\def\C#1{{\bf #1}}

@@

% --- Anders's commands --- %
\newcommand{\ev}{\mathrm{eV}}
\newcommand{\gev}{\mathrm{GeV}}
\newcommand{\mev}{\mathrm{MeV}}
\newcommand{\tev}{\mathrm{TeV}}
\newcommand{\mpc}{m_{\mathrm{p}}c}
\newcommand{\mpcc}{m_{\mathrm{p}}c^2}
\newcommand{\p}{\mathrm{p}}
\newcommand{\pp}{\mathrm{pp}}
\newcommand{\br}{\bold{r}}
\newcommand{\e}{\mathrm{e}}
\newcommand{\ii}{\mathrm{i}}
\newcommand{\inj}{\mathrm{inj}}
\newcommand{\pd}{\partial}
\newcommand{\CR}{\mathrm{CR}}
\newcommand{\gas}{\mathrm{gas}}
\newcommand{\mr}{\mathrm}
\newcommand{\DM}{\mathrm{DM}}
\newcommand{\ISM}{\mathrm{ISM}}
\newcommand{\ICM}{\mathrm{ICM}}
\newcommand{\CMB}{\mathrm{CMB}}
\newcommand{\IC}{\mathrm{IC}}
\newcommand{\mecc}{m_\e c^2}
\newcommand{\mug}{\mathrm{\mu G}}
\newcommand{\rvir}{R_{200}}
\newcommand{\B}{\mathrm{B}}
\newcommand{\ma}{\mathrm{max}}
\newcommand{\low}{\mathrm{low}}
\newcommand{\gae}{\gamma_\mathrm{e}}

@@

The CR electrons cool through inverse Compton (IC) emission and
Coulomb losses on timescales that are relative short compared to the
dynamical timescale of a cluster. We model these looses analytically
by assuming a self similar solution for the CR electron distribution
function \citep{1999ApJ...520..529S}
\begin{eqnarray}
  \label{eq:f_e}
  f_\e(\gamma_\e) &\approx& \frac{f_\rmn{inj,e}(\gamma_\e)}{\alpha-1}
\left(\frac{\gamma_\low}{\gamma_\e}+\frac{\gamma_\e }{\gamma_\ma}\right)^{-1}
\nonumber\\
&\times& \left\{2-\left[\frac{1}{2}\left(1-\frac{\gamma_\e}{\gamma_\ma}+
\left|1-\frac{\gamma_\e}{\gamma_\ma}\right|\right)\right]^{\alpha-1}\right.
\nonumber\\
&-& \left.\left(1+\frac{\gamma_\low}{\gamma_\e}\right)^{-(\alpha-1)}\right\}\,,
\end{eqnarray}
Here $\gamma_\e \approx p_\e = P_\e/m_\e c$ for $ p_\e \gg 1$.  The
approximation in Eqn.~(\ref{eq:f_e}) is valid as long as
$\gamma_\ma\gg \gamma_\low$, where the upper limit is determined from
IC losses through
\begin{eqnarray}
  \label{eq:gamma_max}
b_\IC(\gamma_\e) &=& b_\IC \gamma_\e^2 
= \frac{4}{3}\frac{\sigma_\rmn{T}}{m_\e c} U_\rmn{cmb} \gamma_\e^2 \,,\quad \rmn{and}\nonumber \\
%1.37\times 10^{-20} \gamma^2(1+z)^4 \,\rmn{s}^{-1}
\gamma_\ma &=& \frac{1}{b_\IC\,\Delta t}\,.
\end{eqnarray}
Here $\sigma_\rmn{T}= 8\pi e^4/3(m_\e c^2)^2$ is the Thomson cross
section and $U_\rmn{cmb}$ is the energy density of the CMB in the
cluster. The lower limit is determined from Coulomb losses through
\begin{eqnarray}
  \label{eq:gamma_min}
b_\rmn{C}(\gamma_\e) &=& b_\rmn{C} \gamma_\e^2 
= \frac{3\,\sigma_\rmn{T}\,n_\rmn{el}\,c}{2}
\left[\ln\left(\frac{m_\e c^2 \sqrt{\gamma_\e-1}}{\hbar\,\omega_\rmn{plasma}}\right)\right.\nonumber \\
  &-&\ln(2)\left(\frac{1}{2}+\frac{1}{\gamma_\e}\right)+\frac{1}{2}+
\left.\left(\frac{\gamma_\e-1}{4\gamma_\e}\right)^2\right]
         \,,\quad \rmn{and}\nonumber \\
\gamma_\low &=& b_\rmn{C}\,\Delta t\,.
\end{eqnarray}
Here $\omega_\rmn{plasma} = \sqrt{4\pi e^2 n_\e / m_\e}$ is the plasma
frequency, and $n_\e$ is the number density of free electrons.

We model the final CR electron spectrum as a superposition of five CR
populations, each determined by Eqn.~\ref{eq:f_e} but with a
different spectral index and injected spectra.

\subsection{Synchrotron emission}
WORK UNDER PROGRESS During the merging of two clusters, both
turbulence flows and merging shocks are induced. These processes are
thought to accelerate CRs to high energy, where especially the
distribution function of preexisting CRs is enhanced. In this section
we use the smooth CR electron distribution derived from a simulated
merging cluster in the previous section, boost it using the details of
the shock, and then finally derive the radio synchrotron emission.

We fit the steady state CR electron distribution function in
Fig.~\ref{fig:e_spec} with two power-laws
\begin{equation}
  f_\rmn{e,fit}(\gamma_\e) = C_1\,\gamma_\e^{-\alpha_1}
\theta(\gamma_\rmn{break}-\gamma_\e) 
+ C_2\,\gamma_\e^{-\alpha_2}
\theta(\gamma_\e-\gamma_\rmn{break})\,. 
\label{eq:f_fit}
\end{equation}
The best fit CR electron normalization is given by ${\bf
  C}=(6\times10^{-4},\,10^{7})$, the spectral index by ${\bf
  \alpha}=(3.0,5.5)$, and the energy where the slope changes by
$\gamma_\rmn{break}=1.2\times10^4$.

\C{Probably need to calculate the emissivity numerically since there
  is a break in the power-law spectrum. CHECK} We calculate the radio
synchrotron emissivity of the CR electrons using \cite{1979rpa..book.....R}
\begin{equation}
J_0(\nu) \approx \left(\frac{e^2\,\nu_c}{c}\right)\,\sum_i C_i
\Gamma\left(\frac{3 \alpha_i-1}{12}\right)\,\Gamma\left(\frac{3 \alpha_i+19}{12}\right)\frac{3^\frac{\alpha_i}{2}}{\alpha_i+1}\,\left(\frac{\nu}{\nu_c}\right)^{\frac{1-\alpha}{2}}\,,
\end{equation}
where the cyclotron frequency is given by
$\nu_c=e\,B\,c/(2\pi\mecc)$. We derive the total emitted power from
the source by integrating the emissivity over the target volume. Here,
we assume that the target volume haw an uniform distribution of the
CR electrons, hence the total power per unit frequency is given by
\begin{equation}
  P_0(\nu) = \rmn{Volume} \times J_0(\nu)\,.
\end{equation}

NEED TO REWRITE The typical size of a large relic is of the order of a
Mpc, while the thickness is only of the order 100~kpc. We assume that
the rather unknown extension of the relic along the line of sight to
be about 500~kpc, where the total radio emitting volume is about
$0.05~\rmn{Mpc}^3$. In addition we assume that the average magnetic
field in the cluster outskirts to be of the order
$B\approx0.1\,\mu$G. The total emitted power at 1.4 PHz of the size
typical for a large radio Relic is then given by
$P_0(1.4\,\rmn{PHz})=9\times10^{22}$~W/Hz. This is about a factor
10-100 smaller than what is typically expected from a relic
\cite{2009A&A...508...75V}. However, this luminosity is boosted by the
shock or turbulence induced during a cluster merger. We follow the
prescription outlined in \cite{2011ApJ...734...18K} to calculate the
boost in radio luminosity due to reacceleration, and find that
$P_2(1.4\,\rmn{PHz})\sim\times10^{25}$~W/Hz in agreement with several
of the relics listed in e.g. \cite{2009A&A...508...75V}.

The steady-state test-particle solution of the downstream CR
distribution can be written as a function of the pre-existing CRs through
\begin{equation}
  f_2(\gae) = q \gae^{-q}\int_{\gamma_\rmn{inj}}^\gamma \gamma'^{q-1}
   f_\rmn{e,fit}(\gamma')\dd \gamma'\,.
\end{equation}
Here we have neglected the contribution from the injected CRs. The lowest momentum boundary above which particles can cross the shock is given by 
\begin{equation}
\gamma_\rmn{inj} \approx 1.17 \frac{u_2}{c}\left(1+\frac{1.07}{\epsilon_B}\right)
\left(\frac{M}{3}\right)^{0.1}\,.
\end{equation}

Spectral index
\begin{equation}
q(M) = \frac{3\left(u_1-v_ A\right)}{u_1-v_A-u_2} = 
\frac{3\sigma\left(1-M_A^{-1}\right)}{\sigma-1-\sigma M_A^{-1}}
\end{equation}
where $u_1$ and $u_2$ are the speed of the flow upstream and
downstream, respectively, in the shock rest frame. The compression
ratio is given by $\sigma=u_1/u_2=\rho_2/\rho_1$, and $M_A=u_1/v_A$ is
the upstream Alfvén Mach number with velocity
$v_A=B_1/\sqrt{4\pi\rho_1}$. We derive the test-particle power-law
slope $q$ as a function of the shock Mach number $M$ with
$\sigma=[(\gamma_\rmn{ad}+1)M^2]/[(\gamma_\rmn{ad}-1)M^2+2]$, where
adiabatic index $\gamma_\rmn{ad}=5/3$, $M_A=M c_s/v_A$, and $ c_s$ the
upstream sound speed.

Enhanced emissivity calculated with
\begin{equation}
\frac{J_2(\nu)}{J_0(\nu)} \approx \left(\frac{B_2^{\left(r-1\right)/2}}{B_0^{\left(\alpha-1\right)/2}}\right) \,\left[\frac{f_2(\gae)}{f_0(\gae)}\right]_{\gae=10^4}\,
\left(\frac{\nu}{280\,\rmn{MHz}}\right)^{\left(\alpha-r\right)/2}\,.
\end{equation}

OUTLINE THE PRESCRIPTION eq7, eq21

@@

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
  \includegraphics[width=0.49\columnwidth]{./figures/f_evolution.0.3Rv.eps}
  \includegraphics[width=0.49\columnwidth]{./figures/f_evolution.1.4Rv.eps}
  \caption{Time evolution of cosmic ray (CR) electron spectra. We show
    the CR electron distribution function weighted with the electron
    Lorentz factor ($\gamma_\e$) squared. The time evolution of the CR
    electrons which at $z=0$ reside from (0.3-0.5)$\rvir$ are shown in
    the {\it left panel} and (1.3-1.5)$\rvir$ in the {\it right
      panel}. The black lines show the CR electrons with a
    $\gamma_\e=30$, blue lines $\gamma_\e=200$, and green lines
    $\gamma_\e=10^4$. The solid lines show the CR electrons with full
    cooling, i.e. for Coulomb, inverse Compton, and adiabatic
    losses. Dashed lines show the CR electrons without Coulomb and
    inverse Compton cooling while the dash-dotted lines show the CR
    electrons without adiabatic cooling. The dotted line show the
    injected CR electron distribution function, derived from the
    injected CR protons. For each energy, we normalize the CR electron
    distribution function with the CR electrons without
    cooling. Notice that the high energy part is built up during the
    last Gyr. This is explained by the inverse Compton losses that are
    larger at high energies, especially at high redshifts where the
    energy density of the CMB is much larger than
    today.  \label{fig:e_evol}}
\end{minipage}
\end{figure*}
